---
title: "CooksD Functions"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
      
---

<br><br><br><br>
Reference: Kutner, M., Nachtsheim, C., Neter, J., & Li, W. (2005). Applied linear statistical models (5th ed.). Boston: McGraw-Hill Irwin.  

- Specifically, p. 403: _"For interpreting Cook's distance measure, it has been found useful to relate Di to the F(p, n - p) distribution and ascertain the corresponding percentile value. If the percentile value is less than about 10 or 20 percent, the i th case has little app~t influence on the fitted values. If, on the other hand, the percentile value is near 50 percent or more, the fitted values obtained with and without the i th case should be considered to differ substantially, implying that the ith case has a major influence on the fit of the regression function"_  
- Orig Cook's Distance reference: Cook, R. D., & Weisberg, S. (1982). Residuals and influence in regression. New York: Chapman and Hall.
<br><br><br><br>

# Load packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
```

## Estelle's Original Function
- Inputs for the function will be, for whatever your linear model is looking at 
  - lm.mod.formula <- "DV~IV and covariates"
  - dat.2stage <- dataframe with your data
```{r} 
cooks.distance.comparison <- function(lm.mod.formula, dat.2stage){
  lm1 = lm(lm.mod.formula, dat.2stage)
  par(mfrow = c(2,2))
  plot(lm1)
  
  # Cook's Distance Summary 
  lm1_cook <- cooks.distance(lm1)
  lm1_cook.sum <- matrix(summary(lm1_cook), 
                         ncol=6,
                         dimnames = list(c("Cook's D"), 
                                         c("Min.", "1st Qu.", 
                                           "Median", "Mean", 
                                           "3rd Qu.", "Max")))
  
  # parameters and numobs to use as DoF
  p <- length(coef(lm1))  #number of parameters
  n.mod <- dim(lm1$model)[1] # number of observations
  n_p <- n.mod-p # parameters minus observations 

  # Table of summary stats for cook's D for this model
  tabcook <- as.table(lm1_cook.sum)

  cat("=============================  \ncook's d summary stats")
  print(kable(tabcook, "simple"))
  
  cat("\nMean Cook's D = ")
  cat(c(lm1_cook.sum[4], "  \n"))
  cat("\nTotal N = ")
  cat(c(dim(lm1$model)[1], "  \n"))
  
  cat("\n4/n threshold = ")
  cat(c(4/dim(lm1$model)[1], "  \n"))
  
  cat("\n3*mean threshold = ")
  cat(c(3*mean(lm1_cook), "  \n"))
  
  cat("\nF-Dist 50% = ")
  cat(c(qf(0.5, p, n_p), "  \n"))

  # Identify outliers based on cook's d > 3*mean
  lm1_cook.Out <- lm1_cook > (3*mean(lm1_cook))
  cook.rows <- as.numeric(names(lm1_cook.Out)[lm1_cook.Out == TRUE])
  cook.sub <- dat.2stage$id[cook.rows]
  
  # this makes a dataframe of whichever variables you need, only choosing rows that are identified as above the cook's d threshold specified ([cook.rows])
  # here, you won't need numobs or int.est or slope.est, but would want to grab whichever variables are included and of interest the particular analysis (e.g., dataframe$age[cook.rows])
  cook.threeXmean <- data.frame("ID" = cook.sub, 
                              #"numobs" = dat.2stage$numobsS[cook.rows], 
                              #"int.est" = dat.2stage$int.estS[cook.rows], 
                              #"slope.est" = dat.2stage$slope.estS[cook.rows], 
                              "cooks.d" = lm1_cook[lm1_cook.Out])
  
  
  # Identify outliers based on cook's d > 4/N
  n.mod <- dim(lm1$model)[1]
  lm1_cook.4n <- lm1_cook > 4/n.mod
  cook.rows.4n <- as.numeric(names(lm1_cook.4n)[lm1_cook.4n==TRUE])
  cook.sub.4n <- dat.2stage$id[cook.rows.4n]
  
  cook.4N <- data.frame("ID" = cook.sub.4n, 
                             #"numobs" = dat.2stage$numobsS[cook.rows.4n], 
                              #"int.est" = dat.2stage$int.estS[cook.rows.4n], 
                              #"slope.est" = dat.2stage$slope.estS[cook.rows.4n], 
                              "cooks.d" = lm1_cook[lm1_cook.4n])
  
  # Identify outliers based on cook's d F-distribution percentile ≥ 50%, with p and n-p d.o.f
  f.dist.med <- qf(0.5, p, n_p) #this tells us the 50th percentile of the F-dist with p, n-p d.o.f == cutoff point
  lm1_cook.f <- lm1_cook > f.dist.med
  cook.rows.f <- as.numeric(names(lm1_cook.f)[lm1_cook.f==T])
  cook.sub.f <- dat.2stage$id[cook.rows.f]
  
  cook.fdist <- data.frame("ID" = cook.sub.f,
                           #"numobs" = dat.2stage$numobsS[cook.rows.f],
                           #"int.est" = dat.2stage$int.estS[cook.rows.f],
                           #"slope.est" = dat.2stage$slope.estS[cook.rows.f],
                           "cooks.d" = lm1_cook[lm1_cook.f])
  
  # Identify outliers based on cook's d F-distribution percentile > 20% p and n-p d.o.f
  f.dist.med1 <- qf(0.2, p, n_p)
  lm1_cook.f1 <- lm1_cook > f.dist.med1
  cook.rows.f1 <- as.numeric(names(lm1_cook.f1)[lm1_cook.f1==T])
  cook.sub.f1 <- dat.2stage$id[cook.rows.f1]
  
  cook.fdist1 <- data.frame("ID" = cook.sub.f1,
                            #"numobs" = dat.2stage$numobsS[cook.rows.f1],
                            #"int.est" = dat.2stage$int.estS[cook.rows.f1],
                            #"slope.est" = dat.2stage$slope.estS[cook.rows.f1],
                            "cooks.d" = lm1_cook[lm1_cook.f1])
  
  # plot cook's d with horizontal lines for each threshold
  cols <- c("4/N" = "#3C5488B2", "3*mean" = "#00A087B2", "f-dist > 50%" = "#E64B35B2", "f-dist > 20%" = "#EFC000FF")
  compareplot <- ggplot(lm1, aes(as.numeric(names(cooks.distance(lm1))), cooks.distance(lm1))) + 
    geom_point() + 
    geom_hline(aes(yintercept = (4/n.mod), color = "4/N")) + 
    geom_hline(aes(yintercept = (3*mean(lm1_cook)), color = "3*mean")) + 
    geom_hline(aes(yintercept = f.dist.med, color ="f-dist > 50%")) +
    geom_hline(aes(yintercept = f.dist.med1, color="f-dist > 20%")) + 
    scale_color_manual(name = "threshold", values = cols) + 
    theme(axis.title.x = element_blank()) + labs(title = "Cook's D Threshold Comparison")
  print(compareplot)
  
  # tables for comparison 
  knitr::kables(
    list(knitr::kable(cook.threeXmean, "markdown", caption = "cooks d > 3*mean"),
         knitr::kable(cook.4N, "markdown", caption = "cook's d > 4/N"), "  \n",
         knitr::kable(cook.fdist, "markdown", caption = "f dist > 50%"),
         knitr::kable(cook.fdist1, "markdown", caption = "f dist > 20%")))

}
```


# Ajay's adaptation / functions
- I haven't really explored this completely but it could be helpful to look into because I think it will be more generalizable (maybe email him to clarify what the inputs for each function would be?)
```{r}
###############################################################################
# Check influential outliers based on Cook's distance and different thresholds
# Created by Ajay, November 3, 2020
# Adapted from Estelle's code
###############################################################################

library(tidyverse) 
library(knitr)

print_header <- function(...){
  cat('\n----------------------------------------------------------------------\n')
  cat(...)
  cat('\n----------------------------------------------------------------------\n')
}

print_text <- function(...){
  cat('\n')
  cat(...)
  cat('\n')
}

rounder <- function(num){
  round(num, 4)
}


get_num_influentials <- function(cooks, threshold){
  
  length(which(cooks > threshold))
}

print_influentials <- function(cooks, threshold, data){
  
  print_text('Influential outlier threshold: ', threshold)
  inf.out <- cooks > threshold
  
  # How many data points?
  total <- nrow(data)
  
  # How many influentials?
  num.influentials <- length(which(inf.out))
  percent.influentials <- round((num.influentials/total)*100, 2)
  print_text('Number of influentials: ', num.influentials, '  (', percent.influentials, '%)')   
  
  if(num.influentials > 0) {
    # Which are the influential ids?
    inf.rows <- as.numeric(names(inf.out)[inf.out])
    inf.ids <- as.character(data$ADRCNo[inf.rows])
    
    print_text('Influentials IDs: ')
    print_text(inf.ids)  
    
    # Summary of Influential values
    print_text('Summary of influentials values: ')
    print(summary(cooks[inf.out]))
  }
  
  # Return number of influentials
  num.influentials
}

get_clean <- function(cooks, threshold, modeldata){
  inf.out <- cooks > threshold
  cleaned <- modeldata[!inf.out, ]
  cleaned
}

remove_influentials<- function(model, data){

  # Get data used in the model after removing NAs
  data.mod <- model$model
  
  # Model diagnostics
  par(mfrow = c(2,2))
  plot(model)
  
  # Cook's Distance 
  model.cook <- cooks.distance(model)
  model.cook.summary <- matrix(rounder(summary(model.cook)), 
                         ncol=6,
                         dimnames = list(c("Cook's D"), 
                                         c("Min", "Q1", "Median", "Mean", "Q3", "Max")))
  # parameters and num observations
  n.mod <- nrow(data.mod)
  p <- length(coef(model))
  n_p <- n.mod-p
  
  # Get thresholds
  three.mean <- rounder(3*mean(model.cook))
  four.by.n <- rounder(4/n.mod)
  q.1 <- rounder(qf(0.1, p, n_p))
  q.2 <- rounder(qf(0.2, p, n_p))
  
  # Table of summary stats for cook's D
  tabcook <- as.table(model.cook.summary)
  
  print_header("Cook's D summary stats")
  print(kable(tabcook, "simple"))
  
  print_text('N: ', n.mod)
  print_text('4/n threshold: ', four.by.n)
  print_text('3*Mean threshold: ', three.mean)
  print_text('F-Dist 20% = ', q.2)
  print_text('F-Dist 10% = ', q.1)
  
  
  print_header("Influentials by Cook's D, 4/n threshold")
  print_influentials(model.cook, four.by.n, data)
  
  print_header("Influentials by Cook's D, 3*mean threshold")
  print_influentials(model.cook, three.mean, data)
  
  print_header("Influentials by Cook's D, 20% of F Dist threshold")
  print_influentials(model.cook, q.2, data)
  
  print_header("Influentials by Cook's D, 10% of F Dist threshold")
  print_influentials(model.cook, q.1, data)

  # Plot cook's d with horizontal lines for each threshold
  colors <- c("4/N" = "#3C5488B2", "3*mean" = "#00A087B2", "f-dist > 20%" = "#E64B35B2", "f-dist > 10%" = "#EFC000FF")
  compareplot <- ggplot(model, aes(x = as.numeric(names(model.cook)), y = model.cook)) + 
    geom_point() + 
    geom_hline(aes(yintercept = four.by.n, color = "4/N")) + 
    geom_hline(aes(yintercept = three.mean, color = "3*mean")) + 
    geom_hline(aes(yintercept = q.2, color ="f-dist > 20%")) +
    geom_hline(aes(yintercept = q.1, color="f-dist > 10%")) + 
    scale_color_manual(name = "threshold", values = colors) + 
    theme(axis.title.x = element_blank()) + 
    labs(title = "Cook's D Threshold Comparison",
         y = "Cook's distance")
  print(compareplot)
  
  
  # Finalized threshold
  print_header('Using 10% of F Distribution (', q.1, ') as final threshold')
  final.threshold <- q.1
  num_influentials  <- get_num_influentials(model.cook, final.threshold) 

  # Return data after removing influential outliers (if any) as per desired threshold
  final <- get_clean(model.cook, final.threshold, data.mod)
  
  if(num_influentials > 0) {

    print_text('There were ', num_influentials, ' influential outliers')
    # Update model after removing influential outliers
    model.noout <- lm(formula(model), data = final) 
    
    print_header('Model summary after removing influential outliers')
    print(summary(model.noout))
    
    print_header('Model diagnostics after removing influential points')
    par(mfrow = c(2,2)) 
    plot(model.noout)
  }
    
  # Return clean data if needed
  # final
  
}
```

